---
title: "Getting Started with meandr"
author: "Scott McKenzie"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with meandr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(ggplot2)
library(dplyr)
theme_set(theme_minimal())
set.seed(17)

plot_f <- function(df) {
  ggplot(df, aes(t, f)) +
    geom_point(color = "#175C4A") +
    geom_line(color = "#175C4A")
}
```

## Motivation

When working with time-series data (or any dataset with 1 distinct y-value for each x), there often arises a practical need to develop & optimize a transformation. For example, moving average is used in a wide variety of settings ranging from Daily COVID hospitalizations to financial technical analysis. Such metrics have enormous potential public and/or business implications. This warrants extensive testing prior to implementing a transformation used to generate metric of interest.

Whether it's due to cost or time, the lack of real-world data is a probable challenge in this setting - necessitating the generation of simulated data. However, it might not be feasible or cost-effective to develop a rigorous model fully characterizing the phenomenon of interest. Any coordinates following an arbitrary (yet continuous) trajectory might be sufficient to test and optimize a transformation. `meandr` is designed to fulfill this need. 

## Basic Usage

Each call to `meandr()` outputs coordinates from a **continuously-differentiable-piecewise-polynomial**. In other words: a 2nd-order spline.

```{r}
library(meandr)

example <- meandr() # you can call meandr without explicit arguments
```

```{r echo = FALSE, fig.width=7, fig.height=3}
plot_f(example) +
  labs(title = "example")
```

You can supply arguments if you wish to change default curve behavior. These are discussed in the next section.

```{r eval = FALSE}
meandr(
  n_nodes = 100L,  # number of inflection points (more nodes -> more intricate curve)
  wt = c(1, -1),   # values used to generate 2nd derivative (these are sampled with replacement)
  gain = 0.75,     # controls sensitivity of wt sampling
  n_points = 100L, # number of coordinates (aka resolution)
  scale = 1,       # scales all f(t) values so that max(f(t)) == scale
  seed = NULL      # for reproducibility
)
```

## Arguments

### n_nodes
In general, modulating `n_nodes` should provide sufficient control. Here, the **nodes** correspond to inflection points.

```{r}
nodes_20 <- meandr(n_points = 100,
                   n_nodes = 20,
                   wt = c(1, -1),
                   seed = 2)
```


```{r echo = FALSE, fig.width=7, fig.height=3}
nodes_20 %>% 
  plot_f() +
  labs(title = "nodes_20")
  
```

Holding `seed` constant at `2`, observe that increasing `n_nodes` allows below function to **meander** into new territory. In other words, `df2` is just an augmented variation of `df1`.
```{r}
nodes_100 <- meandr(n_points = 100,
                    n_nodes = 100, # increased from 20
                    wt = c(1, -1),
                    seed = 2)
```


```{r echo = FALSE, fig.width=7, fig.height=3}
nodes_100 %>% 
  mutate(color = t > 0.25) %>% 
  ggplot(aes(t, f)) +
  geom_point(aes(color = color)) +
  geom_line(aes(color = color)) +
  ggsci::scale_color_locuszoom() +
  theme(legend.position = "none") +
geom_vline(xintercept = 0.25, color = "slategray") +
  geom_label(label = "included in nodes_20", fill = "#D43F3AFF", color = "white", x = 0.25, y = 0.35, hjust = 1.05, label.padding = unit(0.4, "lines")) +
  geom_segment(color = "slategray", x = 0.25, xend = 0.2, y = 0.2, yend = 0.2, arrow = arrow(length = unit(0.05, "npc"))) +
  geom_label(label = "new territory with nodes_100", x = 0.25, y = 0.35, hjust = -0.05, label.padding = unit(0.4, "lines"), fill = "#EEA236FF", color = "white") +
  geom_segment(color = "slategray", x = 0.25, xend = 0.3, y = 0.2, yend = 0.2, arrow = arrow(length = unit(0.05, "npc"))) +
  labs(title = "nodes_100")
  
```

### wt & gain

Each call to `meandr` first constructs a piecewise step-function by sampling elements from input `wt`. The resulting step-function serves as the 2nd derivative, $f''(t)$, which `meandr` integrates twice to obtain final result.

This complete procedure is shown below:

```{r eval = FALSE}
meandr(n_points = 200,
       n_nodes = 50,
       wt = c(-1, 1), # default value
       seed = 1010)
```

```{r echo = FALSE, fig.width=7, fig.height=5, echo=FALSE}
set.seed(1010)

nodes <- meandr:::sample_roll(c(1, -1), 50, 0.75)

# define x-values
x <- seq(1/100, 1, length.out = 200)

node_int <- seq(0, 1, length.out = length(nodes) + 1)[-(length(nodes) + 1)]
x_int = rowSums(outer(x, node_int, FUN = meandr:::quasi_greater_equal))
# 2nd derivative
f2 = nodes[x_int]

inc <- c(node_int, 0)[-1] - node_int
c1 <- c(0, cumsum(nodes * inc))
c1 <- c1[-length(c1)]

x_offset <- x - node_int[x_int]
f1 <- x_offset * f2 + c1[x_int]

# parent fn
c0 <- c(0, cumsum(nodes * inc ^ 2 / 2 + c1 * inc))
c0 <- c0[-length(c0)]

f <- f2 * x_offset ^ 2 / 2 + c1[x_int] * x_offset + c0[x_int]

f <- f * 1 / max(abs(f))

gain1 <- tibble(x, `2nd derivative` = f2, `1st derivative` = f1, `final output` = f) %>% 
  tidyr::pivot_longer(-1, names_to = "fn") %>% 
  mutate(fn = forcats::fct_inorder(fn))

gain1 %>% 
  ggplot(aes(x, value, color = fn)) +
  geom_line(size = 1.2) +
  scale_color_manual(values = c("#4EC9DF", "#63A1D0", "#344986")) +
  facet_wrap(~ fn, ncol = 1, scales = "free_y") +
  labs(title = "wt: sampling + integration",
       x = "",
       y = "") +
  theme(legend.position = "none",
        panel.grid.minor = element_blank())
```

The `wt` sampling forming $f''(t)$ utilizes a special algorithm that strives for alternating positive & negative values. This is to prevent `meandr` from generating "boring" curves.

Increasing input `gain` will increase the sensitivity of the `wt` sampling algorithm, ultimately resulting in a curve with greater local variation. Note, however, that the local effects of changing `gain` will cascade with increasing $t$, resulting in an altogether different curve for the same `seed`.

```{r}
gain1 <- meandr(n_points = 200,
                n_nodes = 50,
                gain = 0.75, # default
                seed = 1010)

gain2 <- meandr(n_points = 200,
                n_nodes = 50,
                gain = 2, # increase, this will inject more variance into 2nd derivative
                seed = 1010)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=7, fig.height=3}
gain1 <- mutate(gain1, curve = "gain1")
gain2 <- mutate(gain2, curve = "gain2")

bind_rows(gain1, gain2) %>% 
  ggplot(aes(t, f)) +
  geom_line(aes(color = curve, group = curve), size = 1.2) +
  geom_label(label = "Observe gain2 is 'wigglier'", fill = "#D43F3AFF", color = "white", x = 0.6, y = 0, label.padding = unit(0.25, "lines")) +
  ggsci::scale_color_lancet()
```

The default value of `wt`, `c(1, -1)`, is appropriate for lower levels of `gain` near `0.75`. However, `gain2` curve above looks slightly unnatural, as one can clearly tell this is simply a series of pure quadratic polynomials chained together. We can avoid this effect by adjusting `wt`.

```{r}
custom_wt <- meandr(n_points = 200,
                    n_nodes = 50,
                    wt = seq(-1, 1, by = 0.01), # more variety in 2nd derivative values
                    gain = 2,
                    seed = 1010)
```


```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=7, fig.height=3}
custom_wt %>% 
  ggplot(aes(t, f)) +
  geom_line(color = "#175C4A", size = 1.2) +
  labs(title = "custom_wt")
```

**Beware!** The `wt` algorithm passes `gain` to `exp()`, so you won't observe much difference as $e^{gain} \rightarrow \infty$. I recommend starting at the default value of `0.75` and making incremental adjustments as needed.

```{r eval = FALSE}
gain3 <- meandr(gain = 1000) # 1000 doesn't work because exp(1000) returns Inf

#> Error in sample.int(length(x), size, replace, prob): NA in probability vector
```

### n_points

Use `n_points` to adjust output resolution. This does not affect underlying calculus driving curve shape. Example below uses same seed `2` as before.

```{r}
res1 <- meandr(n_points = 10, # low resolution
               n_nodes = 100,
               seed = 2)

res2 <- meandr(n_points = 1000, # high resolution
               n_nodes = 100,
               seed = 2)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=7, fig.height=3}
library(dplyr)

res1 <- mutate(res1, curve = "res1")
res2 <- mutate(res2, curve = "res2")

bind_rows(res1, res2) %>% 
  ggplot(aes(t, f)) +
  geom_line(aes(color = curve, group = curve), size = 1.2) +
  ggsci::scale_color_lancet() +
  labs(title = "n_points", subtitle = "10 vs 1000")
```

### scale

`meandr` will adjust output so that $max(f) = scale$.

```{r, message = FALSE, warning = FALSE}
library(purrr)

scale <- map_dfr(1:5, ~ {
  meandr(n_points = 100,
         n_nodes = 20,
         wt = c(1, -1),
         scale = .x,
         seed = 20) %>% 
    mutate(scale = .x)
})
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=7, fig.height=3}
scale %>% 
  ggplot(aes(t, f)) +
  geom_line(aes(color = factor(scale), group = factor(scale)), size = 1.2) +
  ggsci::scale_color_locuszoom(name = "scale")
```

This is useful if you want to supplement `meandr` curve with additional noise. In example below, relative amplitude of noise is consistent across different `meandr` curves because we use fixed `scale`.

```{r, eval = FALSE}
library(dplyr)

meandr(scale = 4) %>% 
  mutate(with_noise = f + rnorm(100, sd = 0.2))
```


```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=7, fig.height=7}
library(tidyr)

df1 <- meandr(n_points = 100, n_nodes = 500, wt = c(1, -1), scale = 4, seed = 33)
df2 <- meandr(n_points = 100, n_nodes = 20, wt = c(1, -1), scale = 4, seed = 10)
df3 <- meandr(n_points = 100, n_nodes = 50, wt = c(1, -1), scale = 4, seed = 14)

imap_dfr(list(df1, df2, df3), ~ {
  .x %>% 
    mutate(curve = .y,
           with_noise = f + rnorm(100, sd = 0.2)) %>% 
    rename(without_noise = f) %>% 
    pivot_longer(cols = c(with_noise, without_noise), names_to = "type", values_to = "f")
}) %>% 
  ggplot() +
  geom_line(aes(t, f, color = type, alpha = type), size = 1) +
  scale_alpha_discrete(range = c(1, 0.8)) +
  facet_wrap(~ curve, ncol = 1) +
  scale_color_manual(values = rev(ggsci::pal_lancet()(2))) +
  theme(legend.position = "none",
        panel.spacing = unit(1, "lines"),
        axis.text.x = element_blank()) +
  labs(x = "", y = "")
```
