---
title: "Introduction to meandr"
author: "Scott McKenzie"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to meandr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(ggplot2)
set.seed(17)
theme_set(theme_minimal())
```

## Why meandr?

Suppose we are trying to simulate the behavior of a "somewhat random" time-series phenomenon, $f(t)$.

We could use basic approach below:

```{r, fig.width=7, fig.height=2}
# Set up quick plot function
library(ggplot2)
plot_f <- function(df) {
  ggplot(df, aes(t, f)) +
    geom_point(color = "#175C4A") +
    geom_line(color = "#175C4A")
}

# Generate data
approach_1 <- data.frame(t = 1:100,
                         f = rnorm(100))
plot_f(approach_1)
```

This is okay, but what if we're looking for something with a bit more personality?

Examples:

* Outdoor temperature
* Train station crowd density
* Stock price

Using above `rnorm` approach is not going to adequately illustrate the *character* of these examples.

Instead, we could wrap `rnorm` in `cumsum`:

```{r, fig.width=7, fig.height=2}
approach_2 <- data.frame(t = 1:100,
                         f = cumsum(rnorm(100)))
plot_f(approach_2)
```

This is an improvement, but beyond this, we don't have much control over the signal characteristics. What if we want to generate a smoother curve than above? There's no straightforward way to extend `cumsum(rnorm())` to output a smoother curve while preserving signal resolution. `meandr` offers a solution to this problem.

```{r, fig.width=7, fig.height=3}
library(meandr)

approach_3 <- meandr(n_points = 100,
                     n_nodes = 20,
                     wt = c(1, -1),
                     seed = 10)

plot_f(approach_3)
```

Observe how the curve trajectory in `approach_3` never radically changes between two points. This is the key feature of `meandr`: all data is *continuously differentiable*.

## Usage

### Tuning knobs
The two most important arguments are `n_nodes` (vector) and `wt` (scalar) - adjust these to achieve desired curve behavior.

Increasing `n_nodes` will induce more randomness into curve. However, to the naked eye, resulting curve will smoothen as `n_nodes` exceeds `n_points`.

```{r, fig.width=7, fig.height=3}
plot_f(
  meandr(n_points = 100,
         n_nodes = 100, # increased from 20
         wt = c(1, -1),
         seed = 10)
)
```

On the other hand, small values of `n_nodes` will tend to produce simple parabolic curves.

```{r, fig.width=7, fig.height=3}
plot_f(
  meandr(n_points = 100,
         n_nodes = 3,
         wt = c(1, -1),
         seed = 20)
  )
```
In practice, I recommend using a `wt` of length `2` comprised of positive and negative element (though any numeric vector will be accepted). The default value, `c(1, -1)`, will tend to produce curves with greatest directional variety.

Adjust `wt` values in moderation! As magnitude between `wt[1]` and `wt[2]` deviates, overall curve will veer to +/- `Inf`. Even a slight adjustment, `wt = c(1, -0.8)`, will preduce curves with predominantly positive trend.
```{r, fig.width=7, fig.height=3}
plot_f(
  meandr(n_points = 100,
         n_nodes = 100,
         wt = c(1, -0.8),
         seed = 21
  )
)
```

### Other parameters

Use `n_points` to adjust output resolution. This does not affect underlying calculus which drives curve shape (more details in next section).

```{r}
res1 <- meandr(n_points = 10, # low resolution
               n_nodes = 500,
               wt = c(1, -1),
               seed = 33)

res2 <- meandr(n_points = 1000, # high resolution
               n_nodes = 500,
               wt = c(1, -1),
               seed = 33)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=7, fig.height=3}
library(dplyr)

res1 <- mutate(res1, curve = "res1")
res2 <- mutate(res2, curve = "res2")

bind_rows(res1, res2) %>% 
  ggplot(aes(t, f)) +
  geom_line(aes(color = curve, group = curve), size = 1.2) +
  ggsci::scale_color_lancet()
```

Use `scale` to scale magnitude of y-values. `meandr` will adjust output so that `max(y) = scale`.

```{r, message = FALSE, warning = FALSE}
library(purrr)

scale <- map_dfr(1:5, ~ {
  meandr(n_points = 100,
         n_nodes = 500,
         wt = c(1, -1),
         scale = .x,
         seed = 33) %>% 
    mutate(scale = .x)
})
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=7, fig.height=3}
scale %>% 
  ggplot(aes(t, f)) +
  geom_line(aes(color = factor(scale), group = factor(scale)), size = 1.2) +
  ggsci::scale_color_locuszoom(name = "scale")
```

This is useful if you want to supplement `meandr` curve with additional noise. In example below, relative amplitude of noise is consistent across different `meandr` curves because we use fixed `scale`.

```{r, eval = FALSE}
library(dplyr)

meandr(scale = 4) %>% 
  mutate(with_noise = f + rnorm(100, sd = 0.2))
```


```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=7, fig.height=3}
library(tidyr)

df1 <- meandr(n_points = 100, n_nodes = 500, wt = c(1, -1), scale = 4, seed = 33)
df2 <- meandr(n_points = 100, n_nodes = 20, wt = c(1, -1), scale = 4, seed = 10)
df3 <- meandr(n_points = 100, n_nodes = 50, wt = c(1, -1), scale = 4, seed = 14)

imap_dfr(list(df1, df2, df3), ~ {
  .x %>% 
    mutate(curve = .y,
           with_noise = f + rnorm(100, sd = 0.2)) %>% 
    rename(without_noise = f) %>% 
    pivot_longer(cols = c(with_noise, without_noise), names_to = "type", values_to = "f")
}) %>% 
  ggplot() +
  geom_line(aes(t, f, color = type, alpha = type), size = 1) +
  scale_alpha_discrete(range = c(1, 0.8)) +
  facet_wrap(~ curve, nrow = 1) +
  scale_color_manual(values = rev(ggsci::pal_lancet()(2))) +
  theme(legend.position = "none",
        panel.spacing = unit(1, "lines"),
        axis.text.x = element_blank()) +
  labs(x = "", y = "")
```

## Under the hood

A piecewise step-function is built by sampling elements (with replacement) from input `wt`. This serves as the *2nd derivative*, $f''(t)$, shown below for `approach_3`.

```{r}
wt <- c(1, -1) # default value
set.seed(10)
nodes <- sample(c(1, -1), size = 20, replace = TRUE)
nodes
```
```{r, fig.width=7, fig.height=2, echo=FALSE}
# define x-values
x <- seq(1/100, 1, length.out = 100)

quasi_equal <- function(x, y, tol = 1e-7) {
  abs(x - y) < tol
}
quasi_greater_equal <- function(x, y, tol = 1e-7) {
  quasi_equal(x, y, tol) | (x > y)
}
node_int <- seq(0, 1, length.out = length(nodes) + 1)[-(length(nodes) + 1)]
x_int = rowSums(outer(x, node_int, FUN = "quasi_greater_equal"))
# 2nd derivative
f2 = nodes[x_int]

ggplot(data.frame(t = x, f = f2), aes(t, f)) +
  geom_point() +
  geom_line() +
  labs(y = "f''(t)")
```

Integrating once, we obtain the first derivative, $f'(t)$.

```{r, fig.width=7, fig.height=2, echo=FALSE}
inc <- c(node_int, 0)[-1] - node_int
c1 <- c(0, cumsum(nodes * inc))
c1 <- c1[-length(c1)]

x_offset <- x - node_int[x_int]
f1 <- x_offset * f2 + c1[x_int]

ggplot(data.frame(t = x, f = f1), aes(t, f)) +
  geom_point(color = "#22B6FF") +
  geom_line(color = "#22B6FF") +
  labs(y = "f'(t)")
```

Integrating again, we obtain the final `meandr` curve, $f(t)$.

```{r}
ggplot(approach_3, aes(t, f)) +
  geom_point(color = "#0063A6") +
  geom_line(color = "#0063A6") +
  geom_line(aes(t, f1), data = data.frame(t = x, f = f1), color = "#22B6FF")
```


